---
output:
  html_document:
    highlight: kate
    number_sections: no
    theme: readable
  pdf_document: default
---

# Differetial Expression with DESeq2
## Drosophila melanogaster

### 1. Setup 

First, install [DESeq2](http://bioconductor.org/packages/release/bioc/html/DESeq2.html):

```{r eval=FALSE}
source('http://bioconductor.org/biocLite.R')
biocLite('DESeq2')
```

Then load the libraries we'll need into your R session:

```{r message=FALSE}
library('DESeq2')
library('RColorBrewer')
```

The input files used here can be obtained from /n/ngsdata/workshops/2015_March/fruitfly or
[here](https://software.rc.fas.harvard.edu/ngsdata/2015_March/fruitfly).

Finally, set project.dir to the location of your data files:

```{r}
project.dir <- '~/a/workshops/2015.March/deseq2/fruitfly'
```

---

### 2. Read gene counts into a *data frame*

Read sample gene counts from tab-delimited files into a [data frame](https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.frame.html). The rows of the *data frame* are genes while the columns are samples. 

```{r}
sampleNames <- c('dmel_unf1', 'dmel_unf2', 'dmel_unf3', 'dmel_inf1', 'dmel_inf2', 'dmel_inf3')

fileNames <- unlist( lapply(X = sampleNames, FUN = function(a) paste0(a, '.genes.results')) )
filePaths <- unlist( lapply(X = fileNames, FUN = function(a) file.path(project.dir, a)) )

nsamp = length(sampleNames)

for (i in 1:nsamp) {
    name <- sampleNames[i]
    file <- filePaths[i]
    temp = read.table(file = file, header = TRUE, sep = '\t', row.names = 1)
    if (i == 1) {
        countData = matrix(ncol = nsamp, nrow = nrow(temp), dimnames = list(row.names(temp), rep(NA, nsamp)))
    }
    curNames <- colnames(countData)
    countData[,i] <- as.integer(temp[,'expected_count'])
    curNames[i] <- name
    colnames(countData) <- curNames
}
rm(temp)

countData <- as.data.frame(countData)
dim(countData)
```

Create a second *data frame* for sample information, such as the experimental condition that each sample belongs to:

```{r}
condition <- c('Control', 'Control', 'Control', 'Infected', 'Infected', 'Infected')

colData <- data.frame(row.names=colnames(countData), condition=factor(condition, levels=c('Control','Infected')))

colData
```

---

### 3. Run DESeq 

First, create a DESeqDataSet by specifying the gene counts *data frame*, the sample inforamtion data frame and a design model:

```{r}
dataset <- DESeqDataSetFromMatrix(countData = countData,
                                  colData = colData,
                                  design = ~condition)
dataset
```

Then run the DESeq algorithm and extract results for our two-class comparison:

```{r}
dds <- DESeq(dataset)
result <- results(dds, contrast=c('condition','Infected','Control'))
result <- result[complete.cases(result),]  #remove any rows with NA
head(result)
```

---

### 4. View results

A summary of DESeq results:
```{r}
summary(result)
```

The top 50 up-reguldated and down-regulated genes by p-value:
```{r}
n = 50
resOrdered <- result[order(result$padj),]
topResults <- rbind( resOrdered[ resOrdered[,'log2FoldChange'] > 0, ][1:n,], 
                    resOrdered[ resOrdered[,'log2FoldChange'] < 0, ][n:1,] )
topResults[c(1:5,(2*n-4):(2*n)), c('baseMean','log2FoldChange','padj')]
```

Plot log fold change vs. mean expression for all genes, with genes where p < 0.1 colored red:

```{r}
plotMA(result, main='DESeq2: D. melanogaster Control vs. Infected', ylim=c(-2,2))
```
PCA plot for all genes:
```{r}
rld <- rlogTransformation(dds, blind=TRUE)
plotPCA(rld)
```

Plot counts for a single gene. Below is the plot for the gene with the lowest p-value:
```{r}
plotCounts(dds, gene=which.min(result$padj), intgroup='condition', pch = 19)
```

Display top genes' normalized counts in a heatmap:
```{r}
hmcol <- brewer.pal(11,'RdBu')
nCounts <- counts(dds, normalized=TRUE)
heatmap(as.matrix(nCounts[ row.names(topResults), ]), Rowv = NA, col = hmcol, mar = c(8,2))
```

### 5. Write results to file

```{r}
write.table(result, file = file.path(project.dir,'Control_vs_Infected.tsv'), quote = FALSE, sep = '\t')
```



