---
output:
  html_document:
    highlight: kate
    number_sections: no
    theme: readable
  pdf_document: default
---

# Differential Expression with DESeq2
## Mouse immune cells
## Control vs. treatment samples

### 1. Setup

First, install [DESeq2](http://bioconductor.org/packages/release/bioc/html/DESeq2.html):

```{r eval=FALSE}
source('http://bioconductor.org/biocLite.R')
biocLite('DESeq2')
```

Then load the libraries we'll need into R:

```{r message=FALSE}
library('DESeq2')
library('RColorBrewer')
```

---

### 2. Read gene counts into a *data frame*

Read sample gene counts from a tab-delimited file into a data frame. The rows of the data frame are genes while the columns are samples.

```{r}
countFilePath = 'http://software.rc.fas.harvard.edu/ngsdata/workshops/2015_March/NC11.gene.txt'
countData = read.table(file = countFilePath, header = TRUE, sep = '\t', row.names = 1)
countData = countData[3:ncol(countData)]  #discard chr and strand columns
dim(countData)
```

Read in a second data frame that contains the experimental condition that each sample belongs to:

```{r}
colFilePath = 'http://software.rc.fas.harvard.edu/ngsdata/workshops/2015_March/colData_2conditions.txt'
colData = read.table(file = colFilePath, header = TRUE, sep = '\t', row.names = 1)
colData[['condition']] =  factor(colData[['condition']], levels = c('Control', 'Treatment'))       
colData
```

---

### 3. Run DESeq2

First, create a DESeqDataSet by specifying the gene counts data frame, the sample information data frame and a design model:

```{r}
dataset <- DESeqDataSetFromMatrix(countData = countData,
                                  colData = colData,
                                  design = ~condition)
dataset
```

Then run the DESeq2 algorithm and extract results for our two-class comparison:

```{r}
dds <- DESeq(dataset)
result <- results(dds, contrast=c('condition','Treatment','Control'))
result <- result[complete.cases(result),]  #remove any rows with NA
head(result)
```

---

### 4. View results

A summary of DESeq2 results:
```{r}
summary(result)
```

The top 50 up-regulated and down-regulated genes by p-value:
```{r}
n = 50
resOrdered <- result[order(result$padj),]
topResults <- rbind( resOrdered[ resOrdered[,'log2FoldChange'] > 0, ][1:n,],
                    resOrdered[ resOrdered[,'log2FoldChange'] < 0, ][n:1,] )
topResults[c(1:5,(2*n-4):(2*n)), c('baseMean','log2FoldChange','padj')]
```

Plot log fold change vs. mean expression for all genes, with genes where p < 0.1 colored red:

```{r}
plotMA(result, main=paste0('Condition: Control vs. Treatment'), ylim=c(-5,5))
```

PCA plot for all genes:
```{r}
rld <- rlogTransformation(dds, blind=TRUE)
plotPCA(rld, intgroup = 'condition')
```
Plot counts for a single gene. Below is the plot for the gene with the lowest p-value:
```{r}
plotCounts(dds, gene=which.min(result$padj), intgroup='condition', pch = 19)
```

Display top genes' normalized counts in a heatmap:
```{r}
hmcol <- brewer.pal(11,'RdBu')
nCounts <- counts(dds, normalized=TRUE)
heatmap(as.matrix(nCounts[ row.names(topResults), ]), Rowv = NA, col = hmcol, mar = c(8,2))
```

### 5. Write results to file

```{r}
project.dir <- '~/My_R_Example' 
dir.create(project.dir, showWarnings=FALSE)
write.table(result, file = file.path(project.dir,paste0('NC11_Control_vs_Treatment.tsv')), quote = FALSE, sep = '\t')
```



